---
title: "Class 8: Mini Project"
author: "Katherine Quach (A18541014)"
format: pdf
toc: TRUE
---

## Background

Today we will learn about the methods and technique on clustering and PCA

```{r}
url <- "https://bioboot.github.io/bimm143_W26/class-material/WisconsinCancer.csv"
wisc.df <- read.csv((url))
wisc.data <- read.csv(url, row.names = 1)
```

Have a peak at the first few rows
```{r}
wisc.data <- wisc.data[,-1]
head(wisc.data)
```

Make sure to remove the first `diagnosis` column. I don't want to use this for my machine learning models. We will use it later to compare our results to the expert diagnosis.
```{r}
diagnosis <- wisc.df$diagnosis
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
ncol(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
column_names <- colnames(wisc.data)
length(grep("_mean", column_names))
```

## Principle Component Analysis

We want to scale and center our data before PCA to make sure that each component have equal contribution

The main function here is `prcomp()` and we want to make sure we set the optional 
argument `scale=TRUE`:

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Our main PCA "score plot" or "PC plot" of results:

```{r}
library(ggplot2)
```

```{r}
ggplot(wisc.pr$x) + 
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

```{r}
data.scaled <- scale(wisc.data)
```

Score_plot:

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

The loading vector (wisc.pr$rotation) tells us which original measurements contribute most to each PC.

```{r}
ggplot(wisc.pr$rotation)+
  aes(PC1, 
      reorder(rownames(wisc.pr$rotation), PC1)) +
  geom_col()
```
Collectively these two plots ("score plot" and "loading plot") tell us that if cell nurcleus are deeply indented ("concave"), irregular non circular ("compactedness"), and have large "perimeter" values they tend to be malignant... 

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

## Hierarchical Clustering

First scale the data (`scale`), then calculate a distance matrix (with `dist()` function). Then cluster with the `hclust()` function and plot
```{r}
wisc.hclust <- hclust(dist(scale(wisc.data[,-1])))
plot(wisc.hclust)
abline(h = 20, col="salmon")
```

You can also use `cutree()` function with an argument `k=4` rather than `h=height`

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters)
```

## Combining methods

Here we will take our PCA results and use those as input for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main output from PCA - how the data lie on our new principle component axis/variables) and use a subset of the PCs as input for `hclust()`

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

Cut the dendrogram/tree into two main groups/clusters:

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

I want to know how the cluserin in `grps` with values of 1 or 2 correspond the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

My clustering **Group 1** are mostly "M" diagnosis (179) and my cluster **Group 2** are mostly "B" diagnosis (333)

24 False positive (FP)
179 True positive (TP)
333 True negative (TN)
33 False negative (FN)


```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```
## Sensitivity

TP/(TP+FN)

```{r}
179/(179+33)
```
TN/(TN+FP)

```{r}
333/(333+24)
```

## Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

