---
title: "Class 12 Pt. 1 (RNASeq Galaxy)"
author: "Katherine Quach (A18541014)"
format: pdf
toc: TRUE
---

## Section 1: Proportion of G/G in a population

Downloaded a CSV file from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39895045-39895145;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel>

> Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are
homozygous for the asthma associated SNP (G|G)?

[HINT: You can filter the displayed genotypes by entering the population code MXL. Then either

count those of interest or download a CVS file for this population and use excel or the R functions read.csv(), and table() to answer this question] 

Here we read this CSV file 
```{r}
mxl <-read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head (mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

> Q6. Back on the ENSEMBLE page, use the “search for a sample” field above to find the
particular sample HG00109. This is a male from the GBR population group. What is the
genotype for this sample?

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of GIG

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR populatio than the BKL population.

Let's now dig into this further. 


## Section 4: Population Scale Analysis [OPTIONAL]

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have?

```{r}
expr <- read.table("genotype.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Let's make a box pot

```{r}
ggplot(expr) + 
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```