---
title: "Week 6 Lab Session"
author: "Katherine Quach (A18541014)"
date: "2026-02-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Proportion og G/G in a population

Downloaded a CSV file from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel>

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Let's now dig into this further.


## Section 4: Population Scale Analysis [OPTIONAL]

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have?

```{r}
expr <- read.table("genotype.txt") 
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Let's make a box pot

```{r}
ggplot(expr) + 
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```







