---
title: "Class 05: Data Visualization with GGPLOT"
author: "Katherine Quach (A18541014)"
format: pdf
toc: true
---

## Background

There are multiple ways to make figures in R, and some of these include "Base R" graphics (e.g. `plot()`) and many add-on packages like **ggplot2**

For instance, we will make the same plot with:

```{r}
head(cars)
```

```{r}
plot(cars)
```
First we need to install necessary package with the command 
`install.packages()`

> **N.B.** we never run and install cmd in a quarto code chunk or it will reinstall packages AGAIN, and multiple times (not what we want!)

Every time we want to use one of these "add-on" packages we need to load it up in R with the `library()` function

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs 3 things:

- The **data**, the stuff you want  plotted
- The **aes**thetics, how the data map to the plot
- The **geom**etry, the type of plot

```{r}
ggplot(cars) + 
  aes(x = speed, y = dist) + 
  geom_point()
```

Let's add a line to show the relationship between two entities and then render it out!
```{r}
p <- ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Stopping Distance of Old Cars",
       subtitles = "Data from Cars Object",
       x = "Speed (mph)", y = "Distance (ft)")
p + theme_bw()
```

## Gene expression plot

We can read the input data from the class website
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Let's plot our first version
```{r}
ggplot(genes) + 
  aes(Condition1, Condition2) +
  geom_point()
```

```{r}
table(genes$State)
```

Let's plot version 2 by coloring by `State` so we can see up and down significant genes compared to the unchanging genes
```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col = State) +
  geom_point()
```

Let's plot version 3 by modifying colors to our liking
```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col = State) +
  geom_point() + 
  scale_colour_manual(values = c("skyblue","gray","darkred"))+
  labs(x = "Control(no drug)", y = "Drug", title = "Gene Expression Changes Upon GLP-1") + theme_bw()
```

## Let's go further
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
ggplot(data = gapminder) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point(alpha = 0.3)
```

Let's "facet" (i.e. make a separate plot) by continent
```{r}
ggplot(data = gapminder) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point(alpha = 0.3) + 
  facet_wrap(~continent)
```

## Custom plots

How big is this gapminder dataset?
```{r}
nrow(gapminder)
ncol(gapminder)
```
To filter down a subset of the data, let's use the **dplyr** package

First I need to install it, then load it
`install.packages("dplyr")` and then `library("dplyr")`
```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year == 2007)
head(gapminder_2007)
```

```{r}
filter(gapminder_2007, country == "Ireland")
```

```{r}
filter(gapminder, year == 2007, country == "Ireland")
```

```{r}
filter(gapminder, year == 2007, country == "United States")
```

> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
input <- filter(gapminder, year %in% c(1977,2007))
ggplot(data = input) +
  aes(x = gdpPercap, y = lifeExp, col = continent, size = pop) +
  geom_point(alpha = 0.3) + 
  facet_wrap(~year)
```

## Bar Charts

```{r}
gapminder_top5 <- gapminder %>% 
  filter(year == 2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)
```

> Q. Create a bar chart showing the life expectancy of the five biggest countries by population in 2007

```{r}
ggplot(gapminder_top5) +
  aes(x = reorder(country, -pop), y = pop, fill = country) +
  geom_col(col = "gray30") +
  guides(fill = "none") + labs(x = "Country", y = "Population", title = "Life Expectancy of 5 Biggest Countries in 2007")
```